---
title: "survregVB"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{survregVB}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
csl: apa.csl
---

# Introduction

This vignette demonstrates how to use the `survregVB` function from the
**survregVB** package to fit a log-logistic Accelerated Failure Time
(AFT) model. This function uses Variational Bayes (VB) to infer the
model parameters by maximizing the Evidence Lower Bound (ELBO). We will
use the `rhDNase` dataset from the **survival** package to illustrate
the process. Specifically, we will show you how to prepare the data, set
prior distributions, fit the model, and interpret the results.

## Background

The goal of the `survregVB` function is to fit it a log-logistic AFT
regression model of the following form for right-censored survival time
$T_i$:

$$
log(T_i):=Y=X_i^T+bz_i
$$

where $X_i$ is column vector of $p-1$ covariates and a constant one
(i.e. $X_i=(1,x_i1,...,x_i(p-1))^T$), $\beta$ is a vector of
coefficients, $z_i$ is a random variable following a standard logistic
distribution and $b$ is a scale parameter. In particular, `survregVB`
uses a Bayesian framework to obtain the optimal variational densities of
parameters $\beta$ and $b$ by maximizing the evidence based lower bound
(ELBO) (@jordan1999; @blei2017). In order to do so, we assume prior
distributions:

-   $\beta\sim\text{MVN}(\mu_0,\sigma_0^2I_{p*p})$, with precision
    $v_0=1/\sigma^2$, and

-   $b\sim\text{Inverse-Gamma}(\alpha_0,\omega_0)$

where $\mu_0,v_0,\alpha_0$ and $\omega_0$ are known hyperparameters
(@nedelman2005; @faes2011).

At the end of the model fitting process, the `survregVB` function
obtains the approximated posterior distributions given as follows:

-   $q^*(\beta)$ is a $N_p(\mu,\Sigma)$ density function, and

-   $q^*(b)$ is an $\text{Inverse-Gamma}(\alpha,\omega)$ density
    function,

where the parameters $\mu,\Sigma,\alpha$ and $\omega$ are obtained via
the VB algorithm (@xian2024).

# Installation

First, install the **survregVB** package if not already installed. Since
we'll be using the `rhDNase` dataset, install the **survival** package
as well:

```{r install, eval=FALSE}
#install.packages("survregVB")
#install.packages("survival")
```

# Basic Usage

Once installed, you can load both libraries with the `library()`
function:

```{r setup}
#library(survregVB)
library(survival)
```

# Preparing the Data

The `rhDNase` dataset contains the results of a randomized trial of
rhDNase for the treatment of cystic fibrosis. The survival time, *T*, is
defined as the time until the first pulmonary exacerbation. *T* is
calculated as the difference between the date of entry into the study
(`entry.dt`) and date of last follow-up (`end.dt`), with the follow-up
period capped at 169 days (@survival). The covariates of interest are:

-   Treatment (`trt`: 0=placebo, 1=rhDNase), and
-   Forced expiratory volume (`fev`)

Here is a preview of the `rhDNase` dataset:

```{r}
head(rhDNase)
```

Next, we need to prepare the data for modeling via the example code
provided in the survival package (@survival)::

```{r}
# Extract the first row for each subject
first <- subset(rhDNase, !duplicated(id)) 
dnase <- tmerge(first, first, id=id, tstop=as.numeric(end.dt -entry.dt))

# Subject whose fu ended during the 6 day window after finishing their 
# antibiotics are not counted as new infections
temp.end <- with(rhDNase, pmin(ivstop+6, end.dt-entry.dt))

# Create the event indicator
dnase <- tmerge(dnase, rhDNase, id=id,
                infect=event(ivstart),
                end=  event(temp.end))

# Toss out the non-at-risk intervals, and extra variables
dnase <- subset(dnase, (infect==1 | end==0), c(id:trt, fev:infect))

# Remove duplicated subjects
dnase <- subset(dnase, !duplicated(id))

# Set the survival time 
dnase$time <- dnase$tstop - dnase$tstart

head(dnase)
```

Now, the dataset is ready with the survival time (`time`) and event
indicator (`infect`).

# Defining the AFT Model

Our goal is to fit it a log-logistic AFT regression model of the form:

$${\log(T):=Y=\beta_0+\beta_1x_1+\beta_2x_2+bz}$$ where:

-   $x_1$ is `trt`,

-   $x_2$ is `fev`, and

-   $z$ is a random variable following a standard logistic distribution
    with scale parameter *b (*\@xian2024).

The event of interest is whether or not the subject experienced
infection, `infect`.

## Formula Specification

We define a formula, where the left-hand side is a survival object
returned by the `Surv` function, and the right-hand side consists of the
covariates, `trt` and `infect`. The object returned by the `Surv`
function consists of the time and event of interest (@survival). In this
case, the time is `time` and the event of interest is `infect`:

```{r}
formula <- Surv(time, infect) ~ trt + fev
```

# Setting Prior Distributions

Next we define the prior distributions of the model parameters, *β* and
*b*.

We set hyperparameters $\alpha_0=501$ and $\omega_0=500$ to achieve a
mean scale of one.

```{r}
alpha_0 <- 501
omega_0 <- 500
```

The prior means for $\mu_0$ are chosen based on historical data and
similar analyses on this type of data:

-   for the intercept $\beta_0$, we use $\log(169/2)\approx4.4$ (half
    the follow-up period),

-   for $\beta_1$ (`trt`), we use $\log(1.28)\approx0.25$ (@shah1996)

-   for $\beta_2$ (`fev`) we use $\log(1.04)\approx0.04$ (@block2006)

```{r}
mu_0 <- c(4.4, 0.25, 0.04)
v_0 <- 1
```

# Fitting the model

Now that we have set up the formula and prior distributions, we can fit
the `rhDNase` dataset to the log-logistic AFT model using the
`survregVB` function. The remaining parameters we need to set are:

-   `max_iteration`: The maximum number of iterations the VB algorithm
    will run for with default 100. In this case, we allow up to 10,000
    iterations to ensure that the algorithm has enough time to converge.

-   `threshold`: The convergence threshold based on the change in the
    ELBO between iterations. with default 0.0001 In this case, the
    algorithm will stop if the change in ELBO becomes smaller than
    0.0005.

-   `na.action`: A filter function that specifies how to deal with
    missing data in the provided data-set with default
    `options()$na.action`. In this case, we set it to `na.omit` to omit
    any missing values if present.

```{r}
fit <- survregVB(formula = Surv(time, infect) ~ trt + fev, 
                 data = dnase, 
                 alpha_0 = 501,
                 omega_0 = 500,
                 mu_0 = c(4.4, 0.25, 0.04), 
                 v_0 = 1, 
                 max_iteration = 10000, 
                 threshold = 0.0005,
                 na.action = na.omit)
```

If the VB algorithm reaches the maximum number of iterations before
converging, we will receive a warning message:

```{r}
survregVB(formula = Surv(time, infect) ~ trt + fev, 
          data = dnase, 
          alpha_0 = 501,
          omega_0 = 500,
          mu_0 = c(4.4, 0.25, 0.04), 
          v_0 = 1, 
          max_iteration = 9, 
          threshold = 0.0005)
```

# Interpreting the Results

At the end of the model fitting process, the VB algorithm obtains the
approximated posterior distributions of *b* and *β*:

```{r}
print(fit)
```

## The `survregVB` Object

The `survregVB` function returns a `survreg` object, which is a list
containing the following results:

-   `ELBO`: The value of the converged ELBO. If the algorithm does not
    converge before reaching `max_iteration`, it will be the value of
    ELBO at the final iteration.

-   `alpha`: The final value of parameter $\alpha$ of the approximate
    posterior distribution of *b.*

-   `omega`: The final value of parameter $\omega$ of the approximate
    posterior distribution of *b.*

-   `mu`: The final value of parameter $\mu$ of the approximate
    posterior distribution of *β*.

-   `Sigma`: The final value of parameter $\Sigma$ of the approximate
    posterior distribution of *β*.

-   `iterations`: The number of iterations the algorithm performs before
    converging or reaching `max_iteration`.

-   `n`: The number of observations.

-   `call`: The function call used to invoke the `survregVB` method.

We can access each item in the list using the `$` operator. For example,
if we want to view the evidence lower bound (ELBO):

```{r}
fit$ELBO
```

We can also view summary statistics for the results of the fit via the
`summary` function, which returns the following:

-   `call`: The function call used to invoke the `survregVB` method.

-   `alpha`: The parameter $\alpha$ of the approximate posterior
    distribution of the scale parameter *b.*

-   `omega`: The parameter $\omega$ of the approximate posterior
    distribution of the scale parameter *b.*

-   `mu`: The parameter $\mu$ of the approximate posterior distribution
    of the regression coefficient *β*.

-   `Sigma`: The parameter $\Sigma$ of the approximate posterior
    distribution of the regression coefficient *β*.

-   `iteration`: The number of iterations performed by the \#' VB
    algorithm: before converging or reaching `max_iteration`.

-   `n`: The number of observations.

-   `coefficients`: A matrix with one row for each coefficient, and
    columns containing:

    -   `Value`: The estimated value of each of the regression
        coefficients of *β*.

    -   `Lower CI`: The lower bound of the credible interval associated
        with the coefficient.

    -   `Upper CI`: The upper bound of the credible interval associated
        with the coefficient.

-   `scale`: A vector containing:

    -   `Value`: The estimated value of the scale parameter *b*.

    -   `Lower CI`: The lower bound of the credible interval associated
        with the scale parameter.

    -   `Upper CI`: The upper bound of the credible interval associated
        with the scale parameter.

We can adjust the significance level for the credible intervals by
altering the `ci` argument. The default is set to `0.95`.

```{r}
summary(fit, ci=0.9)
```

# References
